---
title: "EPL Analysis - Team Sports Analytics Buff"
author: "Faizan Husain, Luke Ruan, Akhil Thakur, Neethi Thangiah"
date: "16 September 2021"
output: pdf_document
---
# Intoduction

The English Premier league or the EPL is the top level of the English football league system . The premier league is the most watched sports league in the world, broadcasted to 643 million homes. It generates $4 billion in TV revenue from 2019-2020 alone. It has one of the most match attendance, second only to Bundesliga. Because of such a reputation and financial incentive, its important that we understand various factors affeting the matches. In this markdown, We try to find the factors which influence the most on the teams chance for winning a match.

# Code

The entire code is reproducible and can be accessed from the following github link.

https://github.com/thakurakhil/EPL_ANALYSIS


## Dataset

## Loading the required library

The library "Tidyverse" is essential for the document. Please install it. 


```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# To install tidyverse, please uncomment the below line
# install.packages("tidyverse")
library(tidyverse)

```

## Loading the dataset 

We used the EPL data set sourced from Kaggle. We directly load the data set from the github repository into the data frame.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
url <- "https://raw.githubusercontent.com/thakurakhil/EPL_ANALYSIS/main/Dataset/"

csv_names = c("2000-01.csv","2001-02.csv","2002-03.csv","2003-04.csv","2004-05.csv",
              "2005-06.csv","2006-07.csv","2007-08.csv","2008-09.csv","2009-10.csv",
              "2010-11.csv","2011-12.csv","2012-13.csv","2013-14.csv","2014-15.csv",
              "2015-16.csv","2016-17.csv","2017-18.csv","2018-19.csv","2019-20.csv")

url_csv <- str_c(url , csv_names)

data_all <- url_csv %>% 
              lapply(read_csv, show_col_types = FALSE) %>% bind_rows                                                       


```

```{r}
df <- data_all %>% select("Date","HomeTeam","AwayTeam","FTHG","FTAG","FTR",
                          "HTHG","HTAG","HTR","HST","AST","HF","AF", "HS", 
                          "AS", "Referee","HY", "AY", "HR", "AR","Attendance")
head(as_tibble(df))

```

## Exploring the data set

```{r}
sprintf("The number of rows : %d",dim(df)[1])
sprintf("The number of columns : %d", dim(df)[2])
team_names <- df %>% distinct(HomeTeam)
sprintf("Total Number of Teams : %d", dim(team_names)[1])
```

# Analysis 

## Analysis 1 : Goals Average/ Game through the 20 seasons

```{r}

season_vec <- c("2000-01","2001-02","2002-03","2003-04","2004-05","2005-06",
                "2006-07","2007-08","2008-09","2009-10","2010-11","2011-12",
                "2012-13","2013-14","2014-15","2015-16","2016-17","2017-18",
                "2018-19","2019-20")
season_col <- c()

for(i in 1:20){
  if(i<19){season_col <-  c(season_col,rep(season_vec[i],380))}
  else if( i == 19){season_col <-  c(season_col,rep(season_vec[i],160))}
  else if(i==20){season_col <-  c(season_col,rep(season_vec[i],260))}
}

df$Season <- season_col # Adding it to the date frame df

```
We now calculate the average number of goal scored in the league through the seasons.
```{r}
# Finding the goal average per match/game and summarizing the result in season wise
df_goal_avg_season <- df %>%mutate(TG = FTHG + FTAG) %>%
  group_by(Season) %>%
  summarise(goal_avg = sum(TG)/n())
head(df_goal_avg_season)
```

```{r fig.height=4.5, fig.width=10}
# Plotting the results in graph
ggplot(df_goal_avg_season, aes(Season ,goal_avg ,group = 2)) + 
  geom_point()+ 
  geom_line(col = "red" ,bg = "blue")+
  geom_text(aes(label=signif(goal_avg ,3), size= 3 , vjust = -0.5)) + 
  geom_smooth(method = "lm")+
  ylab("Average Goals per Game")+
  xlab("Seasons")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


## Analysis 1 Inference
We can see an increasing trend in the average number of goals scored from season to season.

## Analysis 2 : Home Ground Advantage 

```{r}

df %>% group_by(FTR) %>% summarize(number_of_wins = n()) %>% mutate(percent = number_of_wins/sum(number_of_wins))
```
```{r}
df_home_ground<-df %>% group_by(HomeTeam,FTR) %>% summarise(wins= n()) %>%
  ungroup() %>% group_by(HomeTeam) %>%  mutate(total_wins = sum(wins)) %>% 
  ungroup() %>% mutate(win_percentage= wins/total_wins) %>% filter(FTR == "H") %>% 
  arrange(desc(win_percentage)) %>% head(11)

boxplot(df_home_ground$win_percentage)
```

```{r fig.height=5, fig.width=10}
ggplot(df %>% group_by(FTR) %>% summarize( number_of_wins = n()) ,
       aes(x = c("Away Team Wins", "Draw", "Home Team Wins") , y = number_of_wins,fill = FTR))+
  geom_bar(stat ="identity") + coord_flip() +theme_grey() +
  geom_text(aes(label=number_of_wins), vjust=0 , hjust = 5 ,size = 5)+
  ylab("Number of Wins")+
  xlab("Full Time Match Results")+
  theme(legend.position = "none")
```

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE}
## Analyzing Home Wins by teams 

ggplot(df_home_ground , aes(x = reorder(HomeTeam ,-win_percentage), y = win_percentage ,fill = HomeTeam)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Teams")+
  ylab("Home Win %")+
  ggtitle( "Home Win Percentage for Teams ")

```

## Analysis 2 Inference 

* Number of wins on home field significantly outnumbers both losses and draws.
* Median winning percentage on home field is >55%.
    + Therefore there seems to be a correlation between playing on home field and a higher win percentage


## Analysis 3 : Do Halftime Leads Translate to Full-time Wins ?
```{r}
## Hypothesis : Result at the end of first half will might well be the final result

df_htr <- cbind(df  %>%filter(HTR == FTR) %>% summarise(HTR_translates_FTR =n()),
                df %>% filter(HTR != FTR) %>% summarize(HTR_doesnt_translate_FTR = n()))
df_htr_pie = data.frame(x <- c(df_htr$HTR_translates_FTR, df_htr$HTR_doesnt_translate_FTR),
                labels <- c("Half Time -> Full Time Translates", "Hall Time -> Full Time Doesn't Translates"))
ggplot(df_htr_pie, aes(x="", y=x, fill=labels)) +geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start=0) +theme_void() + ggtitle("Half time Advantage translation to Full Time") 

df_htr <- df_htr %>% 
  mutate(HTR_translates_FTR_per = HTR_translates_FTR/(HTR_translates_FTR+HTR_doesnt_translate_FTR))
df_htr
```

## Analysis 3 Inference
* Most of the time, winning team at halftime will win the game.
* ~40% of the time, the losing team at halftime will come back to win


## Analysis 4 : Shots Taken vs. Shots On Target
```{r}
df_st_h <- df %>% filter(FTR =='H' ) %>% select(FTHG,HST, HS)
colnames(df_st_h)<- c('GoalsScored','ST', 'S')

df_st_a <- df %>% filter(FTR =='A' ) %>% select(FTAG,AST, AS)
colnames(df_st_a)<- c('GoalsScored','ST', 'S')

df_st_w <- rbind(df_st_a ,df_st_h)
df_st_w$event <- "Winning_Team"

#For Lossing Team

df_st_h <- df %>% filter(FTR =='H' ) %>% select(FTAG,AST, AS)
colnames(df_st_h)<- c('GoalsScored','ST', 'S')

df_st_a <- df %>% filter(FTR =='A' ) %>% select(FTHG,HST, HS)
colnames(df_st_a)<- c('GoalsScored','ST', 'S')

df_st_l <- rbind(df_st_a ,df_st_h)

df_st_l$event <- "Lossing_Team"

df_st <- rbind(df_st_l , df_st_w)

```

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}
#Building a linear model
ggplot(df_st , aes(x= S ,y = ST,colour = event , fill = event)) +
  geom_point(aes(shape=event)) +
  scale_shape_manual(values=c(3,16))+
  geom_smooth(method = "lm")+
  theme_bw() + 
  xlab("Number of Total Shots") +
  ylab("Number of shots on target")
```

## Analysis 4 Inference

* Positive correlation between Shots Taken and Shots On Target.
  + Take more Shots to score more Shots on Target.

## Analysis 5 : Shots On Target vs. Goals Scored

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}
#Building a linear model
ggplot(df_st , aes(x= ST ,y = GoalsScored,colour = event , fill = event)) +
  geom_point(aes(shape=event)) +
  scale_shape_manual(values=c(3,16))+
  geom_smooth(method = "lm")+
  theme_bw() + 
  xlab("Number of shots on target") +
  ylab(" Goals Scored")
```

## Analysis 5 Inference

* Positive correlation between both:
* Shots Taken and Shots On Target
* Shots On Target and Goals Scored
    + Therefore, take more shots!

## Analysis 6 : Does committing more fouls correlate with scoring goals?

```{r}

df_fouls_h <- df %>% filter(FTR =='H' ) %>% select(FTHG,HF,Attendance)
colnames(df_fouls_h)<- c('GoalsScored','Fouls', 'Attendance')

df_fouls_a <- df %>% filter(FTR =='A' ) %>% select(FTAG,AF, Attendance)
colnames(df_fouls_a)<- c('GoalsScored','Fouls', 'Attendance')

df_fouls_w <- rbind(df_fouls_a ,df_fouls_h)
df_fouls_w$event <- "Winning_Team"

#For Lossing Team

df_fouls_h <- df %>% filter(FTR =='H' ) %>% select(FTAG,AF, Attendance)
colnames(df_fouls_h)<- c('GoalsScored','Fouls', 'Attendance')

df_fouls_a <- df %>% filter(FTR =='A' ) %>% select(FTHG,HF, Attendance)
colnames(df_fouls_a)<- c('GoalsScored','Fouls', 'Attendance')

df_fouls_l <- rbind(df_fouls_a ,df_fouls_h)

df_fouls_l$event <- "Lossing_Team"

df_fouls <- rbind(df_fouls_l , df_fouls_w)


```

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(df_fouls , aes(x= Fouls ,y = GoalsScored,colour = event , fill = event)) +
  geom_point(aes(shape=event)) +
  scale_shape_manual(values=c(3,16))+
  geom_smooth(method = "lm")+
  theme_bw() + 
  xlab("Number of Fouls") +
  ylab(" Goals Scored")
```

## Analysis 6 Inference

* Negative correlation between fouls committed and goals scored for the winning team
    + Therefore, commit less fouls

## Analysis 7 : Audience Vs Fouls committed.

Let's see how the audience affect the gameplay. Specifically, lets see if the attendance affects the number of fouls committed by the players.

```{r fig.height=5, fig.width=10, message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(df_fouls , aes(x= Attendance ,y = Fouls,colour = event , fill = event)) +
  geom_point(aes(shape=event)) +
  scale_shape_manual(values=c(3,16))+
  geom_smooth(method = "lm")+
  theme_bw() + 
  xlab("Attendance") +
  ylab("Number of Fouls")
```

## Analysis 7 Inference

* We see that there is no significant affect on the happening of fouls due to the attendance.

# BIAS

* We only looked at EPL. The factors affecting the match might not be significant in other leagues.
* We looked at the entire past 20 of data for out inferences. There might be significant changes in the gameplay in the recent years compared to say last decade. 
* There might be some other factors affecting the gameplay of the match, such as the fouls awarded by the referee. So, let see it that's actually the case.

# Are some referees more inclined to give out more fouls?
```{r}
df_ref <- df %>% group_by(Referee) %>% 
  summarize(count = n(), fouls=sum(HR+AR+AY+HY)) %>% 
  ungroup()

df_ref %>% group_by(Referee) %>%
  ggplot(aes(x = Referee, y = fouls/count)) + 
  geom_bar(stat = "identity") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  xlab("Various Referees") +
  ylab("Fouls Awarded Per Match") +
  ggtitle( "Fouls Awarded by Referees per match") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

# Inference

* We can see that some of the referees award more fouls on average
* So, there is a chance that there might be some bias introduced by the referee.


# Conclusion

# Formula For Victory.

Based on our above analysis, we can conclude that following this formula will lead to victory. 

1. Play on your home field!
2. Get an early lead
3. Take more shots
4. Commit fewer (or smarter) fouls
